---
title: "Powerlifting_Questions"
author: "Joshua Larson"
format: docx
editor: visual
---

# Questions

In this module, we will combine various concepts regarding knowledge in foundational data science, as well as statistical regression techniques.

Here are the needed libraries for the problem set.

```{r}
library(readr)
library(ggplot2)
library(dplyr)
library(tidyr)
library(knitr)
library(forcats)
library(stringr)
```

To start, we need to load the data set.

```{r}
openpowerlifting <- read_csv(here::here("joshlarson/data/openpowerlifting-2025-03-22/openpowerlifting-2025-03-22-c4367451.csv"))
```

This will most likely make your computer look like a 20th century wirebox.

### I. Tidying

1.  Now we need to tidy this data to set our selves up to look at Age vs. Strength, we will worry about class later.

    a\. For [each]{.underline} lifter, keep only tested divisions, only "SBD" events, single-ply or raw equipment. Then get rid of lifters who had a DQ, did not show up, guest lifters, and disqualified for doping (DD). Name this new data set "tidy". Also, keep only the most recent event that lifter competed in by using: 'arrange(desc(Date), .by_group = TRUE) %\>% slice(1)'

    ```{r}

    ```

    b.  Now keep only people who submitted their 'Age', have a `Best3SquatKg` `Best3BenchKg`, and `Best3DeadliftKg`. Because we are only looking at male lifters, set the `Sex` to male. Name this new data set "male"

    ```{r}

    ```

    c.  To create our random sample, we need to select a number from each `AgeClass`. Because of the size of the data set, we can use a large sum of people, but we need to account for the younger and older age classes having less people. I chose 500 as each age classes' sample size. Why is 500 more sufficient than say 20 or 100?

    <!-- -->

    d.  I have supplied the code for generating the random sample. An error that occurred was the age classes were not leveled correctly when we get to the plot. Arrange the `AgeClass` variable accordingly using the `forecats` package. Name all of this "tidy_male"

    ```{r}
    set.seed(1)
    male %>%
      group_by(AgeClass) %>%
      group_modify(~ if (nrow(.x) >= 500) sample_n(.x, 500) else .x) %>%
      ungroup() %>%
      select(1:10, 15, 20, 25:35) %>%
      pivot_longer(
    cols = 11:13,
    names_to = "Lift",
    values_to = "Best_lift"
      ) %>%
      # Arrange the `AgeClass` variable accordingly using the `forecats` package. 
    ```

    e.  Lastly, modify the lift names so they register as "Squat", "Bench", and "Deadlift" using a `str_` function. Test your `readr` knowledge and save this as a csv file named "tidy_male_lifters_quantreg.csv"

    ```{r}

    ```

### II. Modeling and Plotting

In this section, we will explore different plots, identity which models are useful for our investigation, and then apply those models to the plots.

To start, lets plot the points we have:

```{r}
#| fig-height: 14
#| fig-width: 9
tidy_male %>%
  ggplot(aes(x = Age, y = Best_lift)) +
  geom_point(alpha = 0.5, size = 1.2,
            ) + 
  facet_wrap(~ Lift, ncol = 1) + 
  theme_classic() + 
  coord_cartesian(ylim = c(0,NA))
```

2.  If we wanted to add a trend line, we would normally input a `geom_smooth()`.
    a.  What type of smoothing method does `geom_smooth()` default to? Try it out.
    b.  Why is it poor for interpreting performance across age?

What alternative approach could we use to better capture trends in strength as we age? Here, we fit what is known as a polynomial regression model. Why? Because it represents a double exponential relationship.

3.  How does `Age` vs. `Strength` represent a double exponential relationship?

Expanding on what quantile regression is, the model is of the form:

$$Q_Y(\tau \mid X) = \sum_{i=0}^{k} \beta_i(\tau) X^i$$ Where:

-   QY(τ \| X): Estimated τ-th quantile of performance given age X

-   τ: Quantile level (e.g., 0.1 for non-elite, 0.9 for elite)

-   X: Predictor variable (Age)

-   βᵢ: Coefficients specific to each quantile

Quantile regression is used to model the relationship between a predictor,such as `Age`, and different points (quantiles) of the response variable, like `Strength`. Unlike linear regression, which estimates the mean, quantile regression allows us to explore how performance varies across the entire distribution. Different orders will fit the model better, so lets test various orders to see what fits best. To investigate elite vs. non-elite, we will use the 90th and 10th percentiles, or quantiles.

Here we will introduce the `rq()` function from the `quantreg` package and run through some fundamental tips to understand the model. Quantile regression is done using `geom_quantile()` which allows us to model the conditional quantiles of a response variable, not just the mean. This code chunk plots a linear quantile regression line through the median. Determine which factor needs to be changed to get the geom to look at the median

```{r}
library(quantreg)
ggplot(test_male, aes(x = Age, y = TotalKg)) +
  geom_point(alpha = 0.4) +
  geom_quantile(quantiles = 0.5, method = "rq", formula = y ~ x, color = "red", linewidth = 1.2) +
  labs(
    title = "Basic Quantile Regression",
    x = "Age",
    y = "Total Weight Lifted (kg)"
  ) +
  theme_minimal()
```

4.  Here are some questions to get the fundamentals down.
    a.  What is the difference between using `geom_quantile()` vs `geom_smooth()`
    b.  What kind of trend is this line at the moment? Why might this specific trend not be a good representation of `Age` vs. `Strength`?
5.  Below is a code chunk using `geom_quantile()` to create quantile regression plots with a polynomial trend line. In the `formula` argument, replace the "y" inside `poly(x, y)` with a number from 1 to 6 (e.g., `poly(x, 2)` for a 2nd-order polynomial). Based on the plots generated for each order, which polynomial degree provides the best fit for the data, and why? Explain your reasoning using what you observe in the shape and accuracy of the curves.

```{r}
#| fig-height: 14
#| fig-width: 9
tidy_male %>%
  ggplot(., aes(x = Age, y = Best_lift)) +
  geom_point(alpha = 0.5, size = 1.2,
            ) +
  geom_quantile(
    quantiles = c(0.1),
    method = "rq",
    formula = y ~ poly(x, y),
    color = "blue",
    linewidth = 1
  ) +
  geom_quantile(
    quantiles = (0.9),
    method = "rq",
    formula = y ~ poly(x, y),
    color = "green",
    linewidth = 1
  ) + 
  facet_wrap(~ Lift, ncol = 1) + 
  theme_classic() + 
  coord_cartesian(ylim = c(0,NA))
```

6.  Why is looking at trends for different quantiles better than least squares regression which estimates the conditional mean?

Hopefully, now that you have made it here (we are almost done, don't worry), we can now explain the appropriate model for `Age` vs. `Strength`. To model these patterns more accurately, we fit a 4th-degree polynomial to each quantile. This model captures the non-linear relationship between age and total weight lifted, reflecting the rise, peak, and decline of performance over time. The model is: $$
Q_Y(\tau \mid \text{Age}) = \beta_0(\tau) + \beta_1(\tau)\,\text{Age} + \beta_2(\tau)\,\text{Age}^2 + \beta_3(\tau)\,\text{Age}^3 + \beta_4(\tau)\,\text{Age}^4$$

7.  Find the peak age for elite (τ = 0.9) lifters? Non-elite (τ = 0.1)?

8. Using the plot I supplied, construct the critical points on the graph using a vertical, dashed, red line.


### III. Findings and Limitations
Through my study, I concluded that elite lifters develop strength at a faster rate than non-elites, reaching higher totals earlier in their careers. While both groups generally peak between ages 24 and 35, the main difference lies in the developmental phase, where elite lifters show steeper early progress. After the peak, both groups decline at a similar rate, but elite lifters also have a slightly steeper fall of towards the 75-80 age range. These patterns suggest that early performance growth may be a key predictor of long-term success.

However, the current model has limitations. It does not track lifters over time, so we cannot see how individuals improve or decline. Each data point reflects a single competition result from a different athlete. Future work should focus on modeling lifters longitudinally and investigating what specific factors, such as training history, consistency, or recovery, could potentially be the best to predict who becomes elite.